\documentclass{article}

% NeurIPS 2026 style
\usepackage[preprint]{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{subcaption}

\title{CIPHER: A Benchmark for Inverse Transformation Program Recovery in Multimodal Agents}

\author{%
  Maxim Kurkin \\
  \texttt{kurkin@2a2i.org}
}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We introduce \textbf{CIPHER}, a procedurally generated benchmark for
\emph{inverse transformation program recovery} in tool-using multimodal
systems. In CIPHER, a model receives a corrupted video observation together
with natural-language instructions describing the transformations required to
restore it. The system must (i) infer the latent transformation program,
(ii) realize the inverse tool chain via \texttt{ffmpeg}, and
(iii) localize and extract a short target string that is readable only after
successful recovery. This yields a stage-decomposed evaluation of
\textbf{program inference} (S1), \textbf{tool-chain execution} (S2), and
\textbf{terminal-state extraction} (S3), rather than a single monolithic score.

CIPHER spans five calibrated difficulty levels that vary key visibility,
transformation-chain length, instruction ambiguity, and distractor density,
with deterministic seed-based replay and procedural regeneration for
contamination resistance and reproducible regression testing.
We also instantiate a DAVIS-2017 video-temporal split-key domain that measures
multi-frame aggregation and operation inversion with dense process partial
credit.
We evaluate OCR baselines, heuristic pipelines, and frontier multimodal models.
Current systems are strong in S3-only extraction when given the correct
terminal frame, but remain fragile in end-to-end recovery where temporal
localization, transformation sequencing, and executable tool use must all be
correct simultaneously. CIPHER is therefore best viewed as a diagnostic
benchmark for inverse procedural reasoning in multimodal agents, not merely as
another passive video-understanding leaderboard.
\end{abstract}

% ============================================================
\input{sections/01_intro}
\input{sections/02_task}
\input{sections/03_benchmark}
\input{sections/04_experiments}
\input{sections/05_analysis}
\input{sections/06_conclusion}

\section*{Acknowledgements}

\bibliography{references}
\bibliographystyle{plainnat}

\input{checklist}

\end{document}
