\section{Introduction}
\label{sec:intro}

Recent multimodal models are increasingly strong at passive video
understanding: they can answer questions about actions, summarize clips, and
describe temporal events with surprising fluency. But many agentic failures do
not occur at the level of recognizing what is in a video; they occur at the
transition from \emph{understanding} to \emph{executable recovery}. A system
may understand that a video has been mirrored, reversed, or sped up, yet still
fail to realize the correct inverse tool chain, search the right temporal
window, or extract the target signal from the restored result.

CIPHER targets exactly this gap. It instantiates a narrow but revealing
setting: given a corrupted video observation and natural-language
instructions, a system must infer, invert, and execute a latent transformation
program, then localize and extract the target signal. The surface form
resembles a video-editing challenge, but the underlying capability is
\emph{inverse multimodal procedural reasoning} under tool use.

Despite the proliferation of video-capable VLMs~\citep{gpt4o,gemini,qwen2vl},
no benchmark systematically tests this compound capability. While current video
understanding benchmarks such as Video-MME~\citep{videomme} and
LongVideoBench~\citep{longvideobench} assess passive temporal reasoning and
question answering, they do not require active, sequential manipulation of the
visual stream based on language. Conversely, video generation and editing
benchmarks like VBench~\citep{vbench} evaluate perceptual quality rather than
the precise execution of rigid transformational rules. Neither paradigm
captures the tight coupling of instruction-following, procedural tool use, and
visual extraction required here.

Two design choices make CIPHER deliberately interpretable. First, the
transformation vocabulary is intentionally minimal and reversible:
\texttt{reverse}, \texttt{hflip}, \texttt{vflip}, \texttt{rotate90},
\texttt{rotate270}, and \texttt{speed2x}. This keeps failures attributable to
reasoning and execution rather than to an uncontrolled action space. Second,
evaluation is stage-decomposed into program inference (S1), tool-chain
execution (S2), and terminal-state extraction (S3), making it possible to ask
\emph{where} a system fails rather than merely whether the final string is
correct.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{Benchmark for inverse transformation program recovery.}
        We formulate a multimodal task in which the model must recover a latent
        transformation program from corrupted video observations and
        natural-language instructions, then execute the inverse chain and
        extract the target signal.

  \item \textbf{Stage-decomposed evaluation.}
        We decompose the task into program inference (S1), tool-chain
        execution (S2), and terminal-state extraction (S3), allowing
        pipeline-stage attribution instead of a single opaque end-to-end score.

  \item \textbf{Procedural benchmark infrastructure.}
        CIPHER provides deterministic seed-based generation, contamination
        resistance, replayable regression cases, and a calibrated difficulty
        ladder over visibility, chain length, instruction ambiguity, and
        distractor density. We additionally introduce a video-temporal
        split-key domain over DAVIS-2017 clips for fast iteration on temporal
        aggregation and operation inversion.

  \item \textbf{Evidence of an execution gap.}
        Across OCR baselines, heuristic pipelines, and frontier multimodal
        models, strong S3-only extraction does not translate into reliable
        recovery. The current video-temporal slice further shows that many
        failed answers contain meaningful partial progress, exposing a gap
        between perception, temporal evidence selection, and executable policy
        competence.
\end{enumerate}
