\section{The CIPHER Benchmark}
\label{sec:benchmark}

CIPHER is designed as a benchmark for inverse procedural reasoning rather than
as a static collection of adversarial examples. The generator exposes
independent stress axes for policy recovery, while deterministic seeds make
every failure replayable.

\subsection{Difficulty levels}

\begin{table}[h]
\centering\small
\caption{CIPHER difficulty levels (50 samples each, $N{=}250$ total).}
\label{tab:levels}
\begin{tabular}{lccccp{4.5cm}}
\toprule
Level & Key vis. & Ops & NL amb. & Distr. & Primary challenge \\
\midrule
L1 & 1 (large)  & 1 & 1 & 0 & Single-op sanity check \\
L2 & 1 (large)  & 2 & 1 & 0 & Two-op program recovery \\
L3 & 2 (medium) & 2 & 2 & 1 & Paraphrase + distractor pressure \\
L4 & 2 (medium) & 3 & 2 & 2 & Longer chain + distractors \\
L5 & 3 (small)  & 4 & 3 & 3 & Compositional full difficulty \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key visibility}
varies font size (13\%$\to$7\% of frame height), alpha (255$\to$200), and
position randomization (centered$\to$random). The key is always rendered in
bright yellow with a black drop shadow; distractors use non-yellow colors and
are always smaller and dimmer.

\paragraph{Capability axes.}
The benchmark factors difficulty into independent axes that correspond to
distinct stresses on the recovery policy: key visibility controls terminal
extraction difficulty; operation count controls latent-program length;
instruction ambiguity controls inference difficulty; and distractor density
controls false terminal-state attraction. This design supports attribution of
performance drops to specific parts of the policy rather than to a single
undifferentiated ``harder'' setting.

\paragraph{NL ambiguity levels.}
(1) exact canonical phrasing; (2) paraphrase from a pool of templates per
operation; (3) compositional ``undo'' framing with embedded temporal cues. All
phrases are intentionally designed to remain scoreable for S1 baselines.

\paragraph{Operation vocabulary.}
The operational vocabulary consists of six composable transformations:
\texttt{reverse}, \texttt{hflip}, \texttt{vflip}, \texttt{rotate90},
\texttt{rotate270}, and \texttt{speed2x}. These operations are visually
distinct, non-destructive, and composable via standard \texttt{ffmpeg}
filters. The vocabulary is intentionally small to preserve interpretability.

\subsection{Procedural generation and dataset integrity}

Samples are fully deterministic from $(seed, level)$. Background clips are
drawn from a Kinetics-400 pool, and the manifest stores the generated sample
metadata explicitly.

A persistent challenge in evaluating large language and vision models is the
risk of data contamination, where static benchmark data leaks into the
pre-training corpus. CIPHER mitigates this through programmatic dataset
synthesis. The benchmark is generated from $(seed, \text{level})$, which
deterministically selects the background clip, random key, operator chain, and
instruction template. Because samples are constructed on demand using disjoint
background pools, researchers can regenerate the benchmark with novel seeds,
guaranteeing an uncontaminated evaluation environment and a renewable supply of
test instances. The same mechanism also makes failures replayable: a single
seed is sufficient to reconstruct the exact corrupted observation, latent
program, and target timing window for regression-driven debugging.

\subsection{Difficulty calibration via GEPA}
\label{sec:gepa}

A benchmark that either saturates (all models at 100\% EM) or collapses (all
at 0\% EM) provides little discriminative signal. Rather than targeting a
single operating point, we calibrate a ladder of five extraction conditions,
L1--L5, such that a reference frontier VLM spans the full accuracy range from
near-ceiling to near-floor. This design supports informative scaling curves and
helps preserve discriminative power as models improve.

To find the five conditions we adopt a GEPA-style \citep{gepa} reflective
evolutionary loop over the image-mode generator. This search calibrates
\emph{benchmark difficulty}, not model weights. The generator is an
S-expression DSL with tunable numerical parameters including key alpha, font
size, distractor count and color, Gaussian noise, and blur radius. Each
\emph{program} is a fixed instantiation; \emph{fitness} is the squared error
between observed EM and a per-level target profile.

We intentionally run this search first in image mode rather than in the full
video pipeline. Direct video-level calibration mixes extraction difficulty with
pipeline noise from parsing, tool execution, and temporal localization. Image
mode yields a cleaner fitness signal for terminal-state readability, after
which the resulting ladder can be transferred into the full procedural
benchmark.

\begin{table}[h]
\centering\small
\caption{Calibrated extraction levels, measured on Claude Sonnet 4.6
(reference model during calibration, $n{=}8$--16 per level).}
\label{tab:ladder}
\begin{tabular}{llrrr}
\toprule
Level & Config ($\alpha$, font\%, dist, $\sigma$, blur) & target EM & obs.\ EM & char-acc \\
\midrule
L1 & $255$, $12\%$, $0$, $0$, $0$                & $95\%$ & $62\%$ & $94\%$ \\
L2 & $240$, $10.5\%$, $0$, $0$, $0$              & $70\%$ & $69\%$ & $94\%$ \\
L3 & $225$, $9\%$, $0$, $0$, $0$                 & $40\%$ & $44\%$ & $92\%$ \\
L4 & $215$, $8\%$, $1$c, $0$, $0$                & $15\%$ & $25\%$ & $82\%$ \\
L5 & $205$, $7.5\%$, $1$\,(1y), $0.5$, $0$       & $3\%$  & $0\%$  & $55\%$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Metrics}
\label{sec:metrics}

\paragraph{EM.}
$\frac{1}{N}\sum_i \mathbf{1}[\hat{k}_i = k^*_i]$.

\paragraph{CharF1.}
Character-level precision/recall F1 for near-miss credit.

\paragraph{PipeF1.}
\begin{equation}
\text{PipeF1} = 0.25\cdot\overline{\text{S1}} + 0.25\cdot\overline{\text{S2}} + 0.50\cdot\overline{\text{CharF1}}
\end{equation}
S3 receives double weight as the terminal objective; S1 and S2 are diagnostic
leading indicators for policy failure attribution.
