\section{Conclusion}
\label{sec:conclusion}

We introduced CIPHER, a procedurally generated benchmark for inverse
transformation program recovery in tool-using multimodal systems. CIPHER asks
whether a model can infer, invert, and execute a latent transformation program
from corrupted video observations and natural-language instructions, then
localize and extract the target signal. Its three-stage decomposition and
Pipeline-F1 metric enable fine-grained diagnosis of where the recovery policy
breaks down.

Our current results already show a strong asymmetry: frontier VLMs solve S3 key
extraction near-perfectly in isolation, while OCR baselines struggle. The more
important open question, and the central claim CIPHER is designed to measure,
is the gap between this prepared extraction competence and reliable end-to-end
policy execution under temporal search and tool use. The benchmark is therefore
most useful not as a flat leaderboard, but as a diagnostic environment for
stage-attributed failure analysis and reproducible regression testing.

The new DAVIS video-temporal slice strengthens this claim. Even before the full
inverse-\texttt{ffmpeg} table is complete, the split-key domain shows that
strict EM alone hides useful structure: many samples are not solved exactly,
yet model traces contain partially correct operation, frame-localization, or
fragment-normalization evidence. This supports CIPHER's core design principle:
the benchmark should preserve process traces and dense diagnostics rather than
compress all behavior into a single final-string score.

\paragraph{Limitations.}
CIPHER focuses on a small deterministic \texttt{ffmpeg} vocabulary; generative
edits and audio-only keys are out of scope. Human baselines are not included in
this version.

\paragraph{Future work.}
The most important next steps are empirical rather than nominal: complete the
paired S3-only vs full-pipeline table, add scaling curves over chain length,
instruction ambiguity, distractor density, and temporal sparsity, and curate a
canonical failure-seed suite with traces. Beyond that, temporal edit expansion,
multi-frame extraction, and stronger policy-search baselines are natural
extensions.

For the video-temporal domain specifically, the immediate target is a frozen
pool of roughly one hundred audited valid samples, balanced across
\texttt{none}, \texttt{swap\_fragments}, \texttt{reverse\_chars}, and
\texttt{swap\_reverse\_chars}. This pool should be evaluated with
Haiku/Sonnet/Opus, Codex, and Gemini backends under paired RPAD-style debiasing
so that hardness estimates are not dominated by one model family's response
format or tool-use behavior.

\paragraph{Reproducibility.}
All code and generation scripts are released at
\url{https://github.com/Fr0do/cipher}. The benchmark generator is deterministic
by seed and level, enabling replay of individual failure cases.
